---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then:
rmarkdown::render("README.Rmd")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# physiology

<!-- badges: start -->
[![R-CMD-check](https://github.com/jackwasey/physiology/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/jackwasey/physiology/actions/workflows/R-CMD-check.yaml)
[![CRAN status](https://www.r-pkg.org/badges/version/physiology)](https://cran.r-project.org/package=physiology)
[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/jackwasey/physiology?branch=master&svg=true)](https://ci.appveyor.com/project/jackwasey/physiology)
[![Codecov test coverage](https://codecov.io/gh/jackwasey/physiology/branch/master/graph/badge.svg)](https://app.codecov.io/gh/jackwasey/physiology?branch=master)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
<!-- badges: end -->

## Calculate physiologic characteristics of awake and anesthetized adults, children and infants

```{r pkgdesc, results='asis', echo = FALSE}
# This relies on having the package installed
cat(
  unclass(
    utils::packageDescription("physiology")
  )[["Description"]]
)
```

Functions are provided for estimation of a range of physiologic characteristics of infant, child and adult humans.

 - Height, weight and fluid compartments.
 - Henderson-Hasslebalch
 - Alveolar gas equation
 - Airway dead-space
 
Each formula is referenced to the original publication. Warnings can be given for
estimation based on input data outside of normal ranges. Future functions
will cover more material with a focus on anaesthesia, critical
care and peri-operative medicine.

Contributions welcome.
