---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file, then:
rmarkdown::render("README.Rmd")
-->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# physiology

[![Travis build Status](https://travis-ci.org/jackwasey/physiology.svg?branch=master)](https://travis-ci.org/jackwasey/physiology)
[![AppVeyor build status](https://ci.appveyor.com/api/projects/status/github/jackwasey/physiology?branch=master&svg=true)](https://ci.appveyor.com/project/jackwasey/physiology)
[![Coverage status](https://codecov.io/gh/jackwasey/physiology/branch/master/graph/badge.svg)](https://codecov.io/github/jackwasey/physiology?branch=master)
[![CRAN status](https://www.r-pkg.org/badges/version/physiology)](https://cran.r-project.org/package=physiology)
[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)

## Calculate physiologic characteristics of adults and children

```{r pkgdesc, results='asis', echo = FALSE}
cat(utils::packageDescription("physiology")$Description)
```

Functions are provided for estimation of a range of physiologic characteristics of infants, child and adult humans.

 - Height, weight and fluid compartments.
 - Henderson-Hasslebalch
 - Alveolar gas equation
 - Airway dead-space
 
Each formula is referenced to the original publication. Warnings can be given for
estimation based on input data outside of normal ranges. Future functions
will cover more material with a focus on anaesthesia, critical
care and peri-operative medicine.

Contributions welcome.
