---
title: "Effect of breathing circuit compliance and dead-space on neonatal ventilation"
author: "Jack O. Wasey"
date: "October 1, 2017"
output: html_document
vignette: >
  %\VignetteIndexEntry{Effect of breathing circuit compliance and dead-space on neonatal ventilation}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

suppressPackageStartupMessages({
  library(childsds)
  library(dplyr)
  library(magrittr)
})
```

Fully extending the concertina tubing of a breathing circuit reduces compliance: the rigidity of the tubing itself falls, and there is more air in it. The following chart shows the effect of this difference on dead-space fraction in neonatal ventilation. This shows why a compliance test is essential before a neonatal anesthesia case, and why the tubing should not be extended or contracted after this test.

```{r, echo = T}
# units of ml per cmH20
min_compliance <- 0.5
max_compliance <- 1.5
# unit cmH20
driving_pressure <- 15
insp_tube_vol_diff <- driving_pressure * (max_compliance - min_compliance)
```
Get median female weight for age up to one year.
```{r getmedianweights}
dat <- as_tibble(
  childsds::make_percentile_tab(
    childsds::who.ref, 
    item = "weight",
    perc = 50,
    stack = TRUE, 
    sex = "sex"))
dat <- dplyr::filter(dat, sex == "female", age <= 1) %>%
  dplyr::rename(weight = value) %>% 
  dplyr::select(age, weight)
```

```{r deadspacecalc}
dat <- dat %>%
  mutate(tidal_volume = 7 * weight) %>% 
  mutate(error_fraction = insp_tube_vol_diff / tidal_volume)

plot(dat[c("weight", "error_fraction")], type = "l")
```
